### 2021-04-22 final-report.latex/methods.tex
second round of Chris' notes

We implement these two algorithms first in Python for the $\phi^4$ model. Afterwards, we transition to C/C++ code due to the increased speed for more complicated theories.

@@

In the $\phi^4$ model, we can specify a discrete action using the Euclidean action from Eq. \ref{eq:phi4 euclidean action}. We begin by redefining the derivative operator as a difference:
\begin{align}
    \partial_\mu \phi = \frac{\phi\left(x + a \hat{\mu}\right) - \phi(x)}{a}
\end{align}
We can then define the kinetic term
\begin{align}
    \half \left(\partial_t\phi\right)^2 + \half \left(\partial_x\phi\right)^2 \rightarrow& \frac{1}{2a^2}\left[ \left(\phi(x+a \hat{t}) -\phi(x)\right)^2 + \left(\phi(x+a \hat{x}) -\phi(x)\right)^2\right] \\
    \begin{split} \rightarrow& \frac{1}{2a^2}\left[ \phi^2(x+a\hat{t}) + \phi^2(x+a\hat{x}) + 2\phi^2(x)\right.\\ &\qquad \left.- 2\phi(x+a\hat{t})\phi(x) - 2\phi(x+a\hat{x})\phi(x) \right]\hfilneg\end{split}
\end{align}
Since we will eventually sum over all sites $x$, the periodic boundary conditions imply that an overall shift in $x$ does not effect the final action. Therefore, we can combine the first two terms with the third term to produce 
\begin{equation}
    \half \left(\partial_t\phi\right)^2 + \half \left(\partial_x\phi\right)^2 \rightarrow \frac{1}{a^2}\left[ 2\phi^2(x) - \phi(x+a\hat{t})\phi(x) - \phi(x+a\hat{x})\phi(x) \right]
\end{equation}
Unlike the kinetic term, the mass and interaction terms remain unchanged under the discretization procedure. The only remaining change is a shift from an integral to a sum. This takes the form
\begin{equation}
    \label{eq:disc def}
    \int dtdx \rightarrow a^2 \sum_i
\end{equation}
such that the final discretized action becomes
\begin{equation}
    \label{eq:phi4 discretized action}
    S_{\mathrm{lat}}[\phi] = \sum_i \left[-\phi(x_i + a\hat{t})\phi(x_i) - \phi(x_i + a\hat{x})\phi(x_i) + \left(2+\half m_0^2\right)\phi^2(x_i) + \frac{1}{4}\lambda \phi^4(x_i)\right]
\end{equation}

@@

Finally, we redefine the gradient flow in on the lattice. Since the gradient flow is solved exactly in the $\phi^4$ model, we rely on a discrete Fourier Transform. This method isolates the momentum modes of the field and dampens them by a factor of $e^-{\tau p^2}$ where $\tau$ is the flow time and $p$ is the momentum. In the NLSM, the definition of the gradient flow (Eq.~\ref{eq:nsm_gradflow}) becomes 
\begin{equation}
    \label{eq:nsm_gradflow_disc}
    \partial_\tau \e (\tau, x) = \left( 1 - \e(\tau,x) \e(\tau,x)^T\right) \partial^2 \e(\tau,x),
\end{equation}
 where the Laplacian operator $\partial^2$ is defined as
\begin{equation*}
    \partial^2 \e(\tau,x) = \e(\tau, x+a \hat{t}) + \e(\tau,x-a\hat t) + \e(\tau, x+a \hat{x}) + \e(\tau,x-a\hat x) - 4 \e(t)_x.
\end{equation*}
Unlike the $\phi^4$ gradient flow, this differential equation has no analytic solution. Therefore, we numerically solve for the gradient flow using a fourth-order Runge-Kutta approximation (see Sec.~\label{sec:runga-kutta}).

@@

We can also simplify the expression for the magnetic susceptibility. Due to the rotation invariance of both fields, the second terms in Eq.~\ref{eq:chim nlsm} and Eq.~\ref{eq:chim phi4} disappear (see Sec.~\ref{sec:avg mag}). In the NLSM, the susceptibility therefore becomes
\begin{align}
    \chi_m &= L^2 \left\langle \left( \sum_x e(x) \right)^2\right\rangle  \\
           &= L^2 \left\langle \sum_{x,y}\e(x)\cdot \e(y) \right\rangle. 
\end{align}
Due to translational invariance from the periodic boundary conditions, we simplify this expression to be
\begin{equation}
    \chi_m = \left\langle \sum_x \e(0)\cdot\e(x) \right\rangle.
\end{equation}
Following the same logic, we derive 
\begin{equation}
    \chi_m = \left\langle \sum_x \phi(0)\phi(x) \right\rangle
\end{equation}
for the $\phi^4$ model.

@@

Following \cite{berg1981}, this yields an expression for the signed area
\begin{equation}
    A(\e_1, \e_2, \e_1) = 2 \:\mathrm{arg}\Big(1+\e_1\cdot\e_2 + \e_2\cdot\e_3 + \e_3\cdot\e_1 + i\e_1 \cdot (\e_2\times\e_3)\Big).
\end{equation}
Under periodic boundary conditions, these triangles on the sphere necessarily wrap $S^2$ an integer number of times, ensuring $Q$ is an integer. In the continuum limit, the eff\ref{sec:topological charge}
\begin{figure}[h]
\centering
\begin{tikzpicture}
    \pgfdeclarelayer{nodelayer}
    \pgfdeclarelayer{edgelayer}
    \pgfdeclareradialshading{sphere4}{\pgfpoint{-0.2cm}{0.5cm}}% 
        {rgb(0cm)=(1,1,1);
        rgb(1cm)=(0.5,0.5,0.5); rgb(1.05cm)=(1,1,1)}
        %rgb(0.7cm)=(0.1,0.1,0.1); rgb(1cm)=(0.5,0.05,0); rgb(1.05cm)=(1,1,1)}
    \pgfsetlayers{nodelayer,edgelayer}
    \tikzstyle{label}=[fill=none, draw=none, shape=circle]
    \tikzstyle{point}=[inner sep=0pt, minimum size=0.2cm,fill=black, draw, shape=circle]
    %\tikzstyle{interior line}=[{Stealth[scale=1.5]}-,dotted,thick]
    \tikzstyle{interior line}=[dotted,thick]

@@

        \node [style=label] at (8) {\small $A$};
	\end{pgfonlayer}
\end{tikzpicture}
\caption{\label{fig:sphere} Visualization of signed area $A$ on the sphere $S^2$ traced out by field at points $x_1$, $x_2$ and $x_3$.}
\end{figure}
In the NLSM described by Eq.~\ref{eq:nlsm discrete action}, the values of $Q$ are roughly Gaussian around $Q=0$, as shown in Fig.~\ref{fig:hist}.
\begin{figure}[h]
    \centering
      \includegraphics[width=0.6\textwidth]{imgs/hist.png}
      \caption{\label{fig:hist} Histogram of topological charge values $Q$ for trivial NLSM. $L=404$, 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization, $\tau=0$}
\end{figure}

@@

On the lattice, this quantity is known to diverge in the continuum limit \cite{bietenholz2018}. % TODO: highlight that this is true for QCD and the NLSM.

@@

We begin this random walk with a so-called ``hot start'' where each field value at each lattice site is randomly selected. As an alternative to the hot start, we could use a ``cold start'' where the field begins completely aligned. With an appropriate thermalization, these initial configurations have no effect on the Monte Carlo statistics, a fact we verify in Sec.~\ref{thermalization}.

@@

\subsection{Thermalization}
\label{sec:thermalization}
In order to determine the necessary thermalization, we plot the action as a function of Metropolis sweeps in Fig.~\ref{fig:therm}.
\begin{figure}[h]
  \centering
      \begin{subfigure}[b]{0.5\textwidth}\centering
        \includegraphics[width=0.9\textwidth]{imgs/therm24.png}
        \caption{$L=24$}
      \end{subfigure}%
      \hfill
      \begin{subfigure}[b]{0.5\textwidth}\centering
        \includegraphics[width=0.9\textwidth]{imgs/therm404.png}
        \caption{$L=404$}
      \end{subfigure}
      \hfill
      \caption{\label{fig:therm} Plots of the action as a function of Monte Carlo time, starting with a random NLSM lattice.}
\end{figure}
Based on this plot, we determine that 1000 sweeps will give sufficient time for the system to reach the classical action minimum. We use this value for the remainder of this study.

We also compare the hot and cold starts by plotting a histogram of the actions in Fig.~\ref{fig:coldstart}. This plot qualitatively demonstrates the irrelevance of the initial configuration.
\begin{figure}[h]
  \centering
    \includegraphics[width=0.6\textwidth]{imgs/coldstart.png}
  \caption{\label{fig:coldstart} Comparison of cold and hot starts after 1000 sweep thermalization in $L=404$ lattice. 1000 measurements taken every 50 sweeps.}
\end{figure}

>>>>

We implement these two algorithms first in Python for the $\phi^4$ model. Afterwards, we transition to C/C++ code due to the increased speed for more complicated theories. For the C++ simulation, we use the \texttt{gcc} compiler with the highest level of optimization.

@@

In the $\phi^4$ model, we can specify a discrete action using the Euclidean action from Eq. \ref{eq:phi4 euclidean action}. We begin by redefining the derivative operator as a finite difference:
\begin{align}
    \partial_\mu \phi = \frac{\phi\left(x + a \hat{\mu}\right) - \phi(x)}{a}
\end{align}
We can then define the kinetic term
\begin{align}
    \half \left(\partial_t\phi\right)^2 + \half \left(\partial_x\phi\right)^2 \rightarrow& \frac{1}{2a^2}\left[ \left(\phi(x+a \hat{t}) -\phi(x)\right)^2 + \left(\phi(x+a \hat{x}) -\phi(x)\right)^2\right]\nonumber \\
    \begin{split} \rightarrow& \frac{1}{2a^2}\left[ \phi^2(x+a\hat{t}) + \phi^2(x+a\hat{x}) + 2\phi^2(x)\right.\\ &\qquad \left.- 2\phi(x+a\hat{t})\phi(x) - 2\phi(x+a\hat{x})\phi(x) \right]\hfilneg\end{split}
\end{align}
Since we will eventually sum over all sites $x$, the periodic boundary conditions imply that an overall shift in $x$ does not effect the final action. Therefore, we can combine the first two terms with the third term to produce 
\begin{equation}
    \half \left(\partial_t\phi\right)^2 + \half \left(\partial_x\phi\right)^2 \rightarrow \frac{1}{a^2}\left[ 2\phi^2(x) - \phi(x+a\hat{t})\phi(x) - \phi(x+a\hat{x})\phi(x) \right]
\end{equation}
Unlike the kinetic term, the mass and interaction terms remain unchanged under the discretization procedure. The only remaining change is a shift from an integral to a sum. This takes the form
\begin{equation}
    \label{eq:disc def}
    \int dtdx \rightarrow a^2 \sum_i
\end{equation}
such that the final discretized action becomes
\begin{equation}
    \label{eq:phi4 discretized action}
    S_{\mathrm{lat}}[\phi] = \sum_i \left[-\phi(x_i + a\hat{t})\phi(x_i) - \phi(x_i + a\hat{x})\phi(x_i) + \left(2+\half m_0^2\right)\phi^2(x_i) + \frac{1}{4}\lambda \phi^4(x_i)\right]
\end{equation}

@@

Finally, we redefine the gradient flow in on the lattice. Since the gradient flow is solved exactly in the $\phi^4$ model, we rely on a discrete Fourier Transform. This method isolates the momentum modes of the field and dampens them by a factor of $e^{-\tau p^2}$ where $\tau$ is the flow time and $p$ is the momentum. In the NLSM, the definition of the gradient flow (Eq.~\ref{eq:nsm_gradflow}) becomes 
\begin{equation}
    \label{eq:nsm_gradflow_disc}
    \partial_\tau \e (\tau, x) = \left( 1 - \e(\tau,x) \e(\tau,x)^T\right) \partial^2 \e(\tau,x),
\end{equation}
 where the Laplacian operator $\partial^2$ is defined as
\begin{equation*}
    \partial^2 \e(\tau,x) = \e(\tau, x+a \hat{t}) + \e(\tau,x-a\hat t) + \e(\tau, x+a \hat{x}) + \e(\tau,x-a\hat x) - 4 \e(t)_x.
\end{equation*}
Unlike the $\phi^4$ gradient flow, this differential equation has no analytic solution. Therefore, we numerically solve for the gradient flow using a fourth-order Runge-Kutta approximation (see Sec.~\ref{sec:runga-kutta}).

@@

We can also simplify the expression for the magnetic susceptibility. Due to the spin-flip invariance of both fields, the second terms in Eq.~\ref{eq:chim nlsm} and Eq.~\ref{eq:chim phi4} disappear (see Sec.~\ref{sec:avg mag}). In the NLSM, the susceptibility therefore becomes
\begin{align}
    \chi_m &= L^2 \left\langle \left( \sum_x e(x) \right)^2\right\rangle  \\
           &= L^2 \left\langle \sum_{x,y}\e(x)\cdot \e(y) \right\rangle. 
\end{align}
Due to translational invariance from the periodic boundary conditions, we simplify this expression to be
\begin{equation}
    \chi_m = \left\langle \sum_x \e(0)\cdot\e(x) \right\rangle.
\end{equation}
Following the same logic, we derive 
\begin{equation}
    \chi_m = \left\langle \sum_x \phi(0)\phi(x) \right\rangle
\end{equation}
for the $\phi^4$ model.

@@

Following \cite{berg1981}, this yields an expression for the signed area
\begin{equation}
    A(\e_1, \e_2, \e_1) = 2 \:\mathrm{arg}\Big(1+\e_1\cdot\e_2 + \e_2\cdot\e_3 + \e_3\cdot\e_1 + i\e_1 \cdot (\e_2\times\e_3)\Big).
\end{equation}
Under periodic boundary conditions, these triangles on the sphere necessarily wrap $S^2$ an integer number of times, ensuring $Q$ is an integer. In the continuum limit, field must be uniform at $x\rightarrow\infty$ and therefore effectively forms a sphere (see Sec.~\ref{sec:topological charge}).
\begin{figure}[h]
\centering
\begin{tikzpicture}
    \pgfdeclarelayer{nodelayer}
    \pgfdeclarelayer{edgelayer}
    \pgfdeclareradialshading{sphere4}{\pgfpoint{-0.2cm}{0.5cm}}% 
        {rgb(0cm)=(1,1,1);
        rgb(1cm)=(0.5,0.5,0.5); rgb(1.05cm)=(1,1,1)}
        %rgb(0.7cm)=(0.1,0.1,0.1); rgb(1cm)=(0.5,0.05,0); rgb(1.05cm)=(1,1,1)}
    \pgfsetlayers{nodelayer,edgelayer}
    \tikzstyle{label}=[fill=none, draw=none, shape=circle]
    \tikzstyle{point}=[inner sep=0pt, minimum size=0.2cm,fill=black, draw, shape=circle]
    %\tikzstyle{interior line}=[{Stealth[scale=1.5]}-,dotted,thick]
    \tikzstyle{interior line}=[dotted,thick]

@@

        \node [style=label] at (8) {\small $A$};
	\end{pgfonlayer}
\end{tikzpicture}
\caption{\label{fig:sphere} Visualization of signed area $A$ on the sphere $S^2$ traced out by field at points $x_1$, $x_2$ and $x_3$.}
\end{figure}
In the NLSM described by Eq.~\ref{eq:nlsm discretized action}, the values of $Q$ are roughly Gaussian around $Q=0$, as shown in Fig.~\ref{fig:hist}.
\begin{figure}[h]
    \centering
      \includegraphics[width=0.6\textwidth]{imgs/hist.png}
      \caption{\label{fig:hist} Histogram of topological charge values $Q$ for trivial NLSM. $L=404$, 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization, $\tau=0$}
\end{figure}

@@

On the lattice, this quantity is known to diverge in the continuum limit owing solely to the $x=0$ term \cite{bietenholz2018}. This divergence exists in QCD as well \cite{bruno2014}.

@@

We begin this random walk with a so-called ``hot start'' where each field value at each lattice site is randomly selected. As an alternative to the hot start, we could use a ``cold start'' where the field begins completely aligned. With an appropriate thermalization, these initial configurations have no effect on the Monte Carlo statistics, a fact we verify in Sec.~\ref{sec:thermalization}.

@@

\subsection{Thermalization}
\label{sec:thermalization}
In order to determine the necessary thermalization, we plot the action as a function of Metropolis sweeps in Fig.~\ref{fig:therm}.
\begin{figure}[h]
  \centering
      \begin{subfigure}[b]{0.5\textwidth}\centering
        \includegraphics[width=\textwidth]{imgs/therm24.png}
        \caption{$L=24$}
      \end{subfigure}%
      \hfill
      \begin{subfigure}[b]{0.5\textwidth}\centering
        \includegraphics[width=\textwidth]{imgs/therm404.png}
        \caption{$L=404$}
      \end{subfigure}
      \hfill
      \caption{\label{fig:therm} Plots of the action as a function of Monte Carlo time, starting with a random NLSM lattice.}
\end{figure}
Based on this plot, we determine that 1000 sweeps will give sufficient time for the system to reach the classical action minimum. We use this value for the remainder of this study.

We also compare the hot and cold starts by plotting a histogram of the actions in Fig.~\ref{fig:coldstart}. This plot qualitatively demonstrates the irrelevance of the initial configuration.
\begin{figure}[h]
  \centering
    \includegraphics[width=0.6\textwidth]{imgs/coldstart.png}
  \caption{\label{fig:coldstart} Histogram of lattice-averaged actions $S/L^2$ with hot and cold starts. 1,000 sweep thermalization in $L=404$ lattice, 1,000 measurements taken every 50 sweeps.}
\end{figure}

### 2021-04-22 final-report.latex/results.tex
fix Chris' notes in Methods
second round of Chris' notes

We initially implement the $\phi^4$ model to verify the results of our system. According to previous studies \cite{monahan2016, schaich2006}, the $\phi^4$ model exhibits a symmetric and broken phase depending on its parameters $m_0^2$ and $\lambda$, specifically occurring at $m_0^2 = -0.72$ when $\lambda = 0.5$. We verify this result by plotting four observables: the lattice average $|\langle\bar\phi\rangle|$, the magnetic susceptibility $\chi$, the Binder cumulant $U$ and the bimodality $B$ in Fig.~\ref{fig:phi4}.
\begin{figure}[h]
    \centering
      \includegraphics[width=0.7\textwidth]{imgs/phi4.png}
      \caption{\label{fig:phi4} The lattice average $|\langle\bar\phi\rangle|$, the lattice susceptibility $\chi$, the Binder cumulant $U$ and the bimodality $B$ plotted as functions of $m_0^2$. $L=32$, $\lambda=0.5$. The lattice was thermalized from a hot start for 1000 sweeps. Afterwards, 100 measurements were taken with 100 sweeps between each. The red horizontal line indicates $U=2/3$, the symmetric limit of the Binder cumulant.}
\end{figure}
This figure confirms a phase transition near $m_0^2=-0.72$ and supports the accuracy of our model.

@@

\subsection{Comparison with Existing Literature}
Following \cite{berg1981}, we approximate the internal energy in the strong ($\beta<1$) and weak ($\beta>2$) regimes as 
\begin{equation}
    \label{eq:analytic_energy}
E \approx \begin{cases} 
    4-4y-8y^3-\frac{48}{5}y^5& \beta<1 \\
    \frac{2}{\beta} + \frac{4\beta^2} + 0.156\frac{1}{\beta^3}& \beta>2. 
\end{cases}
\end{equation}
We compare this analytical result and simulated values of $\chi_m$ with the Monte Carlo simulation in Fig.~\ref{fig:bergluscher}.
\begin{figure}[h]
    \centering
      \includegraphics[width=0.6\textwidth]{imgs/internal_energy.png}
      \caption{\label{fig:bergluscher}Comparison with \cite{berg1981}. First panel: internal energy compared with analytic  energy (Eq.~\ref{eq:analytic_energy}). Second panel: magnetic susceptibility compared with literature values.}
\end{figure}
These two charts show a high degree of agreement with the current literature.

We also seek to confirm the results from Bietenholz et al.\cite{bietenholz2018}. Specifically, that the topological susceptibility $\chi_t$ diverges in the continuum limit even at finite flow time. Since $\chi_t$ is in units of inverse distance squared, we multiply by $\xi_2^2$, the square of the second moment correlation length, to achieve a scale-invariant value $\chi_t\xi_2^2$. Additionally, we use a parameter $t_0$ to scale the flow time such that $t_0\sim L^2$. In our Monte Carlo simulation, we adapt the same values as \cite{bietenholz2018} for $\xi_2$.

To begin the comparison, we plot $\chi_t\xi_2^2$ as a function of flow time $\tau$, shown in Fig.~\ref{fig:bietenholz}.
\begin{figure}[h]
    \centering
      \includegraphics[width=\textwidth]{imgs/bietenholz.png}
      \caption{\label{fig:bietenholz} $\chi_t L^2$ as a  function of flow time $\tau$. Simulation run with 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization.}
\end{figure}
We find that the flow time effectively decreases the topological susceptibility by dampening high-momentum modes. To analyze the divergence of $\chi_t$ in the continuum limit, we plot $\chi_t \xi_2^2$ as a function of lattice size $L$. We perform this simulation at flow time $\tau=5t_0$.
\begin{figure}[h]
    \begin{center}
      \begin{subfigure}[b]{\textwidth}
          \includegraphics[width=\textwidth]{imgs/divergence.png}
          \caption{$\tau = 0$}
      \end{subfigure}

      \begin{subfigure}[b]{\textwidth}
          \includegraphics[width=\textwidth]{imgs/divergence_flowed.png}
          \caption{$\tau = 5t_0$}
      \end{subfigure}
      \caption{\label{fig:bietenholz} $\chi_t\xi_2^2$ as a function of $L$. The data is fit with both a log and power fit. Simulation run with 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization. In the $\tau=0$ case, we have compared our result with the curve fit found in \cite{bietenholz2018}.}
    \end{center}
\end{figure}
The data is fit with two options: a log fit
\begin{equation}
    \chi_t \xi_2^2 = A \mathrm{log}(B L) + C
\end{equation}
and a power fit
\begin{equation}
    \chi_t \xi_2^2 = A B^L + C.
\end{equation}
The data fits these functions with $\chi^2/DOF$ of $31.3$ and $761.6$ respectively. Both of these function diverge as $L\rightarrow \infty$, indicating that the topological susceptibility does as well. This is still a slower divergence than the linear divergence featured in the $\tau=0$ case.

\subsection{Topological charge when $\theta\neq 0$}
Following the method explained in Sec.~\ref{sec:topotheta}, we calculate the imaginary part of $\langle Q \rangle$ for arbitrary $\theta$. We perform this calculation for four values of the flow time $\tau$, shown in Fig~\ref{fig:theta}.
\begin{figure}[h]
    \centering
      \includegraphics[width=\textwidth]{imgs/theta.png}
      \caption{\label{fig:theta} Imaginary part of $\langle Q \rangle$ as a function of $\theta$. Simulation run with 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization. Note the different scaling of the $y$-axis.}
\end{figure}
These plots demonstrate the divergence of the continuum limit in the $\tau=0$ and the flowed regimes. In the $\tau=0$ case, the slope increases sharply, reflecting the linear divergence of $\chi_t$. However in the flowed regime, this divergence is much slower, reflecting the logarithmic or power-law relationship shown in Fig.~\ref{fig:bietenholz}.

%\chapter{Statistical Analysis}

>>>>

Using the Monte Carlo procedure outlined in Sec.~\ref{sec:methods}, we run statistical simulations on the $\phi^4$ model and NLSM.
\section{$\phi^4$ results}
We initially implement the $\phi^4$ model to verify the results of our system. According to previous studies \cite{chang1976, monahan2016, schaich2006}, the $\phi^4$ model exhibits a symmetric and broken phase depending on its parameters $m_0^2$ and $\lambda$, specifically occurring at $m_0^2 = -0.72$ when $\lambda = 0.5$. We verify this result by plotting four observables: the lattice average $|\langle\bar\phi\rangle|$, the magnetic susceptibility $\chi$, the Binder cumulant $U$ and the bimodality $B$ in Fig.~\ref{fig:phi4}.
\begin{figure}[h!]
    \centering
      \includegraphics[width=0.9\textwidth]{imgs/phi4.png}
      \caption{\label{fig:phi4} The lattice average $|\langle\bar\phi\rangle|$, the lattice susceptibility $\chi$, the Binder cumulant $U$ and the bimodality $B$ plotted as functions of $m_0^2$. $L=64$, $\lambda=0.5$. The lattice was thermalized from a hot start for 1000 sweeps. Afterwards, 1000 measurements were taken with 50 sweeps between each. The red horizontal line indicates $U=2/3$, the symmetric limit of the Binder cumulant.}
\end{figure}
This figure confirms a phase transition near $m_0^2=-0.72$ and supports the accuracy of our model.

@@

\subsection{Comparison with Existing Literature}
Following \cite{berg1981}, we approximate the internal energy in the strong ($\beta<1$) and weak ($\beta>2$) regimes as 
\begin{equation}
    \label{eq:analytic_energy}
E \approx \begin{cases} 
    4-4y-8y^3-\frac{48}{5}y^5& \beta<1 \\
\frac{2}{\beta} + \frac{4}{\beta^2} + 0.156\frac{1}{\beta^3}& \beta>2
\end{cases}
\end{equation}
where 
\begin{equation}
    y=\mathrm{coth}\beta-\frac{1}{\beta}.
\end{equation}
We compare this analytical result and simulated values of $\chi_m$ with the Monte Carlo simulation in Fig.~\ref{fig:bergluscher}.
\begin{figure}[h!]
    \centering
      \includegraphics[width=0.6\textwidth]{imgs/internal_energy.png}
      \caption{\label{fig:bergluscher}Comparison with \cite{berg1981}. First panel: internal energy compared with analytic  energy (Eq.~\ref{eq:analytic_energy}). Second panel: magnetic susceptibility compared with literature values.}
\end{figure}
These two charts show a high degree of agreement with the current literature however there is a slight discrepancy, prehaps arising from different Monte Carlo methods.

We also seek to confirm the results from Bietenholz et al. \cite{bietenholz2018}. Specifically, we show the topological susceptibility $\chi_t$ diverges in the continuum limit even at finite flow time. Since $\chi_t$ is in units of inverse distance squared, we multiply by $\xi_2^2$, the square of the second moment correlation length, to achieve a scale-invariant value $\chi_t\xi_2^2$. Additionally, we use a parameter $t_0$ to scale the flow time such that $t_0\sim L^2$. In our Monte Carlo simulation, we adapt the same values as \cite{bietenholz2018} for $\xi_2$.

To begin the comparison, we plot $\chi_t\xi_2^2$ as a function of flow time $\tau$, shown in Fig.~\ref{fig:bietenholz}.
\begin{figure}[h!]
    \centering
      \includegraphics[width=\textwidth]{imgs/bietenholz.png}
      \caption{\label{fig:bietenholz} $\chi_t L^2$ as a  function of flow time $\tau$. Simulation run with 10,000 measurements every 50 sweeps, 1,000 sweep thermalization.}
\end{figure}
We find that the flow time effectively decreases the topological susceptibility by dampening high-momentum modes. To analyze the divergence of $\chi_t$ in the continuum limit, we plot $\chi_t \xi_2^2$ as a function of lattice size $L$. We perform this simulation at flow times $\tau=0$ and $\tau=5t_0$ (Fig.~\ref{fig:divergence}).
\begin{figure}[h!]
    \begin{center}
      \begin{subfigure}[b]{\textwidth}
          \centering
          \includegraphics[width=0.8\textwidth]{imgs/divergence.png}
          \caption{$\tau = 0$}
      \end{subfigure}

      \begin{subfigure}[b]{\textwidth}
          \centering
          \includegraphics[width=0.8\textwidth]{imgs/divergence_flowed.png}
          \caption{$\tau = 5t_0$}
      \end{subfigure}
      \caption{\label{fig:divergence} $\chi_t\xi_2^2$ as a function of $L$. The data is fit with both a logarithmic and power fit. Simulation run with 10,000 measurements, measurements every 50 sweeps, 1,000 sweep thermalization. In the $\tau=0$ case, we have compared our result with the curve fit found in \cite{bietenholz2018}.}
    \end{center}
\end{figure}
The data is fit with two options: a log fit
\begin{equation}
    \chi_t \xi_2^2 = a \mathrm{log}(b L + c)
\end{equation}
and a power law fit
\begin{equation}
    \chi_t \xi_2^2 = A L^B + C.
\end{equation}
We calculate the parameters to these functions using the \texttt{curve\_fit} tool in the \texttt{scipy} Python package \cite{virtanen2020}. When $\tau=5t_0$, the data fits these functions with $\chi^2/DOF$ of \logchi and \powchi respectively, indicating errors were underestimated. Both of these function diverge as $L\rightarrow \infty$, indicating that the topological susceptibility does as well in the continuum limit. Though there is a clear difference between the quantitative fit from \cite{bietenholz2018} and the fit calculated in this work, both demonstrate divergent behavior. This result supports the inherent divergence of $\chi_t$ in the continuum limit.

\subsection{Topological charge when $\theta\neq 0$}
Following the method explained in Sec.~\ref{sec:topotheta}, we calculate the imaginary part of $\langle Q \rangle$ for arbitrary $\theta$. We perform this calculation for four values of the flow time $\tau$, shown in Fig~\ref{fig:theta}.
\begin{figure}[h!]
    \centering
      \includegraphics[width=\textwidth]{imgs/theta.png}
      \caption{\label{fig:theta} Imaginary part of $\langle Q \rangle$ as a function of $\theta$. Simulation run with 10,000 measurements, measurements very 50 sweeps, 1,000 sweep thermalization. Note the different scaling of the $y$-axis.}
\end{figure}
These plots demonstrate the divergence of the continuum limit in the $\tau=0$ and the flowed regimes. In the $\tau=0$ case, the slope increases sharply, reflecting the linear divergence of $\chi_t$. However in the flowed regime, this divergence is much slower, reflecting the logarithmic or power-law relationship shown in Fig.~\ref{fig:bietenholz}.

\section{Implications}

Berg \& L\"uscher \cite{berg1981} originally illuminated this discrepancy between the renormalization group hypothesis (that $\chi_t \rightarrow 0$) and the numerical results, providing three possible causes: 
\begin{enumerate}
    \item The definition of the topological charge does not scale to the continuum.
    \item There are ultraviolet divergences.
    \item There is no reasonable continuum limit.
\end{enumerate}
Since the gradient flow suppresses ultraviolet fluctuations, the persistence of a divergent topological susceptibility under the gradient flow undermines the second option, thereby supporting the other two.

### 2021-04-22 final-report.latex/theory.tex
second round of Chris' notes

\subsection{Non-linear sigma model}

The non-linear sigma model (NLSM) is a prototypical theory for a variety of unrelated physical phenomena, including applications in string theory \cite{callan1985} and ferromagnetism \cite{polyakov1975}. As a simple nonperturbative model, it also provides an ideal starting point for lattice QCD studies. Specifically, the NLSM exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability.

>>>>

\subsection{Non-linear sigma model}
The non-linear sigma model (NLSM) is a prototypical theory for a variety of unrelated physical phenomena, including applications in string theory \cite{callan1985} and ferromagnetism \cite{polyakov1975}. As a simple nonperturbative model, it also provides an ideal starting point for lattice QCD studies. Specifically, the NLSM exhibits many properties shared by Yang-Mills gauge theories, such as a mass gap, asymptotic freedom and $O(2)$ renormalizability.
